---
title: "Exercice 5"
author: "Mikael Jumppanen"
output: html_document
---


# Exercice 5. Dimensionality reduction techniques


## 2. Read data and brief description  


```{r}
human <- read.table("human.csv", row.names = 1, sep=";", header = T, stringsAsFactors = F, dec = ",")

# look at the (column) names of human
names(human)

# look at the structure of human
str(human)

dim(human)
# print out summaries of the variables

summary(human)
```


**The dataset combines several indicators from most countries in the world**


**Brief description of the variables**      

- **GNI** = Gross National Income per capita        
- **Life.Exp** = Life expectancy at birth       
- **Edu.Exp** = Expected years of schooling        
- **Mat.Mor** = Maternal mortality ratio        
- **Ado.Birth** = Adolescent birth rate        
- **Parli.F** = Percetange of female representatives in parliament

>Variables used to derive variables(**not included**):      

- **Edu2.F** = Proportion of females with at least secondary education 
- **Edu2.M** = Proportion of males with at least secondary education    
- **Labo.F** = Proportion of females in the labour force      
- **Labo.M** = Proportion of males in the labour force       

> Derived variables(included):    

- **Edu2.FM** = Edu2.F / Edu2.M        
- **Labo.FM** = Labo2.F / Labo2.M 


## Garphical overview of the data     

```{r}
library(GGally)
library(corrplot)
library(dplyr)

# visualize the 'human_' variables
ggpairs(human)

# compute the correlation matrix and visualize it with corrplot
cor(human) %>% corrplot()
```

Data seems to be quite well normally distributed except GNI is quite a lot at left side. It seems also that there is two distributions for maternal moralility? Maybe developping countries have their own distribution. 

It seems that maternal mortality ration correlates with women education and expected years of schooling positively with birth rate. 

Longer life is also expected if women education is increased and education generally increased.(or at least it correlates doesn't tell anything about causality)    

# PCA non standardized data and standardized data

```{r}
# print out summaries of the variables
summary(human)

# perform principal component analysis (with the SVD method)
pca_human <- prcomp(human)

# draw a biplot of the principal component representation and the original variables
biplot(pca_human, choices = 1:2, cex=c(0.8,1), col=c("grey40", "deeppink2"))
```

```{r}

# standardize the variables
human_std <- scale(human)

# print out summaries of the standardized variables
summary(human_std)

# perform principal component analysis (with the SVD method)
pca_human <- prcomp(human_std)

# draw a biplot of the principal component representation and the original variables
biplot(pca_human, choices = 1:2, cex=c(0.8,1), col=c("grey40", "deeppink2"))
```

Results are quite different after standadization because in the first plot variables with biggest values have biggest effect. With standardization each variable contributes equally to variation.

```{r}
# create and print out a summary of pca_human
s <- summary(pca_human)
print(s)

# rounded percetanges of variance captured by each PC
pca_pr <- round(100*s$importance[2, ], digits = 1)

# print out the percentages of variance

print(pca_pr)
# create object pc_lab to be used as axis labels
pc_lab <- paste0(names(pca_pr), " (", pca_pr, "%)")

# draw a biplot
biplot(pca_human, cex = c(0.8, 1), col = c("grey40", "deeppink2"), xlab = paste(pc_lab[1],"General Health"), ylab = paste(pc_lab[2], "Women education"))
```

## Peronal interpretation       

Variables related to general health explain 50% of the variation of the first principal component and 16.4 % of the variation is exlained by women education in the second principal component.         


```{r}
library("FactoMineR")
library("tidyr")
library("ggplot2")
data(tea)
# column names to keep in the dataset
keep_columns <- c("Tea", "How", "how", "sugar", "where", "lunch")

# select the 'keep_columns' to create a new dataset
tea_time <- select(tea, one_of(keep_columns))

# look at the summaries and structure of the data
str(tea_time)
summary(tea_time)

# visualize the dataset
gather(tea_time) %>% ggplot(aes(value)) + facet_wrap("key", scales = "free")  + geom_bar() + theme(axis.text.x = element_text(angle=45, hjust=1, size=8))


# multiple correspondence analysis
mca <- MCA(tea_time, graph = FALSE)

# summary of the model

summary(mca)

# visualize MCA
plot(mca, invisible=c("ind"), habillage="quali")
```

                
It seems that tea shops sell unpackaged tea and chain stores tea bags. Black tea is usually drank without sugar and earl grey with sugar and milk. 
